Predicting Electoral Outcomes: Bayesian Modeling of Voter Support Trends for Kamala Harris and Donald Trump in the 2024 Presidential Election

### 2024-11-04 paper/paper.qmd
delete some words for prose

---
title: "Predicting Electoral Outcomes: Bayesian Modeling of Voter Support Trends for Kamala Harris and Donald Trump in the 2024 Presidential Election"
subtitle: "Predictions shows higher likelyhood for Harris to be the Winner"
author: 
  - Xinqi Yue
  - Yawen Tan
  - Duanyi Su
thanks: "Code and data are available at: [https://github.com/xinqiyue/2024_US_Election_Prediction](https://github.com/xinqiyue/2024_US_Election_Prediction)."
date: today
date-format: long
abstract: "This paper builds a predictive Bayesian model to forecast the 2024 US presidential election using a “poll-of-polls” approach, analyzing high-quality polls for Kamala Harris and Donald Trump. The model incorporates data on polling methodology, state trends, and candidate support. Our findings indicate that although no candidate is guaranteed to win a total of 270 electoral votes due to lack of polls in certain states, Kamala Harris is more likely to win the election with 222 electoral votes against Donald Trump's 109 electoral votes. This study contributes to an understanding of electoral forecasting, helping us interpret aggregated polling data to anticipate election results."
format: pdf
number-sections: true
table-of-contents: true
toc-depth: 2
bibliography: references.bib
---

@@

# Introduction
The 2024 U.S. presidential election is generating significant public interest, with recent polling reflecting shifts in candidate support across various demographic and regional groups. Polling data, while insightful, presents challenges due to inherent biases, variability in polling methodology, and regional influences on voter behavior. This study aims to address these challenges using a "poll-of-polls" approach, aggregating high-quality national and state polls to forecast voter support for Kamala Harris and Donald Trump. By applying Bayesian modeling, this analysis incorporates both the temporal dynamics of candidate support and state-specific trends.

Our model predicts trends in voter support for the 2024 U.S. Presidential Election, focusing on Donald Trump and Kamala Harris. The results, visualized in @fig-modelvisual, indicate stable support for Harris at an estimated 52% nationally, with a confidence interval of approximately 49% to 55%. Trump’s predicted support level centers around 46%, with a range of about 43% to 49%, showing a slight downward trend as the election date nears. State-specific polling results highlight variances, with Harris showing stronger support in traditionally Democratic areas like Maine CD-1, where predicted support reaches around 62%. In contrast, battleground states such as Georgia, Florida, and Pennsylvania exhibit closely contested support levels, reflecting competitive races in these regions. These predictions underscore the variability in voter sentiment across states, illustrating the importance of both national and regional factors shaping the electoral landscape.

In remainder of this paper, we structure our analysis as follows. @sec-data provides a detailed overview of the data used, including the data cleaning process, explanations of the predictor and outcome variables, as well as other variables that contribute to the model. We also present relevant data visualizations to illustrate initial trends and patterns. @sec-model focuses on the model setup and methodology, specifically detailing the Bayesian model used, its underlying assumptions, and software choices. @sec-result presents our main results, including both the model's outputs and the predictions of voter support for the key candidates. In @sec-discussion, we discuss the findings, offer insights into the broader implications, and highlight the limitations and potential directions for future research. Finally, the appendices provide additional methodological analysis, an idealized survey methodology, and supporting details for the data and model diagnostics.

## Estimand
Our main goal is to predict who will win the election, donald trump or Kamala Harris. Specifically, we estimated the voter support rate percentage of Donald Trump and Kamala Harris on election day considering sample size, state and end date, and then calculated the total number of votes obtained by Trump and Harris through the total number of votes in each state, so as to provide a reliable prediction of the election results.

@@

## Data Overview {#sec-data-overview}
@sec-data centers on a polling dataset for Donald Trump and Kamala Harris, filtered for quality by including only results from pollsters with a numeric grade above 2.5. Key variables such as state, end date, sample size, candidate name, and support percentage (pct) are identified. The data cleaning process standardizes column names, retains relevant fields, and addresses missing values to ensure high-quality data for analysis. The section includes a table presenting the first six entries of the cleaned analysis dataset, alongside multiple graphs: a bar graph comparing candidate support percentages by state, scatter plots illustrating trends in candidate support over time, and visualizations examining the relationship between sample size and support percentages.

The analysis utilizes FiveThirtyEight's dataset of national presidential general election polls [@fivethirtyeight]. Following the approach outlined in [@tellingstories], we aim to predict the election outcome based on this polling data. The analyses were conducted in R [@citeR], utilizing several packages to streamline the process. The tidyverse packages [@citetidyverse] were essential for data simulation and preliminary testing. Initially, the raw data was downloaded using the arrow package [@citearrow], which facilitated efficient handling of large datasets.

@@

- **state**: The U.S. state where the poll was conducted, which allows for state-by-state analysis and comparison of support levels. @fig-state indicates the distribution of states.The histogram shows the distribution of polling data across different states. States like California and Georgia have high counts, indicating overrepresentation, while states such as Wyoming and South Dakota are underrepresented.

@@

In this analysis, we utilize a Bayesian linear regression model to predict the percentage of voter support for Donald Trump and Kamala Harris in the 2024 U.S. Presidential Election. This choice of modeling framework is particularly suitable for our objectives, as it allows for the incorporation of prior knowledge regarding electoral dynamics while updating predictions based on newly observed data. The Bayesian approach provides a robust mechanism for estimating uncertainty in our predictions, which is crucial in the context of political polling where public sentiment can fluctuate significantly.
The model structure includes key predictors: the polling end date, candidate name, sample size, and state. The use of these predictors is justified by their established relevance in previous electoral analyses, highlighting their roles in shaping voter preferences. The inclusion of `end_date` facilitates the examination of temporal trends in support, reflecting how candidate visibility and public sentiment evolve as the election date approaches.
We model the outcome variable, support percentage (pct), as a normally distributed random variable influenced by these predictors. The mathematical formulation allows for the representation of the intercept, which denotes the baseline support level, alongside coefficients that quantify the influence of each predictor. The priors assigned to the model parameters are carefully chosen to reflect reasonable expectations based on historical voting patterns, thus enhancing the model's interpretability and effectiveness.
Moreover, the Bayesian framework enables the modeling of potential interactions between state-specific effects and candidate support, allowing us to capture the regional variations that are often observed in electoral outcomes. By incorporating individual pollster effects as fixed coefficients, we can account for the inherent biases associated with different polling organizations, ensuring that our model provides a nuanced view of candidate support.
This model's formulation strikes a balance between complexity and interpretability, allowing for a comprehensive analysis of the factors influencing voter sentiment while maintaining computational efficiency. The Bayesian approach not only enhances our understanding of electoral dynamics but also provides a statistically sound basis for forecasting potential election outcomes. Overall, the chosen model is well-aligned with the objectives of this study, facilitating a robust analysis of polling data to derive meaningful insights into the upcoming election.

>>>>

---
title: "Predicting Electoral Outcomes: Bayesian Modeling of Voter Support Trends for Kamala Harris and Donald Trump in the 2024 Presidential Election"
subtitle: "Predictions shows higher likelyhood for Harris to be the Winner"
author: 
  - Xinqi Yue
  - Yawen Tan
  - Duanyi Su
thanks: "Code and data are available at: [https://github.com/xinqiyue/2024_US_Election_Prediction](https://github.com/xinqiyue/2024_US_Election_Prediction)."
date: today
date-format: long
abstract: "This paper builds a predictive Bayesian model to forecast the 2024 US presidential election using a poll-of-polls approach, analyzing high-quality polls for Kamala Harris and Donald Trump. The model incorporates data on polling methodology, state trends, and candidate support. Our findings indicate that although no candidate is guaranteed to win a total of 270 electoral votes due to lack of polls in certain states, Kamala Harris is more likely to win the election with 222 electoral votes against Donald Trump's 109 electoral votes. This study contributes to an understanding of electoral forecasting, helping us interpret aggregated polling data to anticipate election results."
format: pdf
number-sections: true
table-of-contents: true
toc-depth: 2
bibliography: references.bib
---

@@

# Introduction
The 2024 U.S. presidential election is generating significant public interest, with recent polling reflecting shifts in candidate support across various demographic and regional groups. Polling data presents challenges due to inherent biases, variability in polling methodology, and regional influences on voter behavior. This study aims to address these challenges using a poll-of-polls approach, aggregating high-quality national and state polls to forecast voter support for Kamala Harris and Donald Trump. By applying Bayesian modeling, this analysis incorporates both the temporal dynamics of candidate support and state-specific trends.

Our model predicts trends in voter support for the 2024 U.S. Presidential Election, focusing on Donald Trump and Kamala Harris. The results, visualized in @fig-modelvisual, indicate stable support for Harris at an estimated 52% nationally, with a confidence interval of approximately 49% to 55%. Trump’s predicted support level centers around 46%, with a range of about 43% to 49%, showing a decreasing trend as the election date nears. State-specific polling results highlight variances, with Harris showing stronger support in traditionally Democratic areas like Maine CD-1, where predicted support reaches around 62%. In contrast, battleground states such as Georgia, Florida, and Pennsylvania exhibit closely contested support levels, reflecting competitive races in these regions. These predictions underscore the variability in voter sentiment across states, illustrating the importance of both national and regional factors shaping the electoral landscape.

In remainder of this paper, we structure our analysis as follows. @sec-data provides a detailed overview of the data used, including the data cleaning process, explanations of the predictor and outcome variables, as well as other variables that contribute to the model. We also present relevant data visualizations to illustrate initial trends and patterns. @sec-model focuses on the model setup and methodology, specifically detailing the Bayesian model used, its underlying assumptions, and software choices. @sec-result presents our main results, including both the model's outputs and the predictions of voter support for the key candidates. In @sec-discussion, we discuss the findings, give broader implications, and highlight the limitations and potential directions for future research. Finally, the appendices provide additional methodological analysis, an idealized survey methodology, and model diagnostics.

## Estimand
Our main goal is to predict who will win the election, Donald trump or Kamala Harris. Specifically, we estimated the voter support rate percentage of Donald Trump and Kamala Harris on election day considering sample size, state and end date, and then calculated the total number of votes obtained by Trump and Harris through the total number of votes in each state, so as to provide a reliable prediction of the election results.

@@

## Data Overview {#sec-data-overview}
@sec-data centers on a polling dataset for Donald Trump and Kamala Harris, filtered for quality by including only results from pollsters with a numeric grade above 2.5. Key variables such as state, end date, sample size, candidate name, and support percentage (pct) are identified. The data cleaning process standardizes column names, retains relevant fields, and addresses missing values to ensure high-quality data for analysis. The section includes a table presenting the first six entries of the cleaned analysis data set, alongside multiple graphs: a bar graph comparing candidate support percentages by state, scatter plots illustrating trends in candidate support over time, and visualizations examining the relationship between sample size and support percentages.

The analysis utilizes FiveThirtyEight's data set of national presidential general election polls [@fivethirtyeight]. Following the approach outlined in [@tellingstories], we aim to predict the election outcome based on this polling data. The analyses were conducted in R [@citeR], utilizing several packages to streamline the process. The tidyverse packages [@citetidyverse] were essential for data simulation and preliminary testing. Initially, the raw data was downloaded using the arrow package [@citearrow], which facilitated efficient handling of large datasets.

@@

- **state**: The U.S. state where the poll was conducted, which allows for state-by-state analysis and comparison of support levels. @fig-state indicates the distribution of states.The histogram shows the distribution of polling data across different states. States like California and Georgia have high counts, indicating over representation, while states such as Wyoming and South Dakota are underrepresented.

@@

In this analysis, we utilize a Bayesian linear regression model to predict the percentage of voter support for Donald Trump and Kamala Harris in the 2024 U.S. Presidential Election. This choice of modeling framework is particularly suitable for our objectives, as it allows for the incorporation of prior knowledge regarding electoral dynamics while updating predictions based on newly observed data. The Bayesian approach provides a robust mechanism for estimating uncertainty in our predictions, which is crucial in the context of political polling where public sentiment can fluctuate significantly.
The model structure includes key predictors: the polling end date, candidate name, sample size, and state. The use of these predictors is justified by their established relevance in previous electoral analyses, highlighting their roles in shaping voter preferences. The inclusion of `end_date` facilitates the examination of temporal trends in support, reflecting how candidate visibility and public sentiment evolve as the election date approaches.
We model the outcome variable, support percentage (pct), as a normally distributed random variable influenced by these predictors. The mathematical formulation allows for the representation of the intercept, which denotes the baseline support level, alongside coefficients that quantify the influence of each predictor. The priors assigned to the model parameters are carefully chosen to reflect reasonable expectations based on historical voting patterns, thus enhancing the model's interpret ability and effectiveness.
Moreover, the Bayesian framework enables the modeling of potential interactions between state-specific effects and candidate support, allowing us to capture the regional variations that are often observed in electoral outcomes. By incorporating individual pollster effects as fixed coefficients, we can account for the inherent biases associated with different polling organizations, ensuring that our model provides a nuanced view of candidate support.
This model's formulation strikes a balance between complexity and interpret ability, allowing for a comprehensive analysis of the factors influencing voter sentiment while maintaining computational efficiency. The Bayesian approach not only enhances our understanding of electoral dynamics but also provides a statistically sound basis for forecasting potential election outcomes. Overall, the chosen model is well-aligned with the objectives of this study, facilitating a robust analysis of polling data to derive meaningful insights into the upcoming election.
